---
title: "Data Wrangling and Cleaning: Steps 1-11"
author: "David Mullaly"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## Introduction

This document covers the data wrangling process and all eleven data cleaning steps for the problem. We load multiple relational tables, merge them into a single flat file, and perform comprehensive data cleaning to prepare for missing data analysis.

### Data Cleaning Steps Covered:
**Steps 1-5 (Initial Cleaning)**

1. Open data in your software of choice
2. Review variables for common sense based on SME knowledge
3. Review how the software coded the variables (nominal, continuous)
4. Perform data integrity/validation checks
5. Handle dates

**Steps 6-7 (Categorical & Zero-Variance)**

6. Handle categorical variables - keep as is, combine rare levels, combine similar levels
7. Handle zero-variance predictors

**Steps 8-11 (Advanced Cleaning)**

8. Handle near zero-variance predictors
9. Eliminate redundant columns and linear combination columns
10. Search for outliers and initial search for missing values
11. Sanity check using Decision Tree (1 to 2 splits)

```{r setup, include=FALSE}
# Setup Chunk - Load Required Libraries
library(mice)
library(VIM)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(corrplot)
library(lubridate)
library(flextable)
library(rpart)       # Decision Tree package
library(rpart.plot)  # Visualization of Decision Trees

setwd("C:/Users/david/Desktop/Spring-Semester-/Missing Data")
```

## Data Loading and Wrangling

Load four relational tables and merge into a single flat file (FFdf) using left joins on Cust_ID.

```{r load_data, echo=FALSE}
# Load the raw data files
MainDF <- read.csv("Raw.csv")
StoreDF <- read.csv("StoreTable.csv")
ConcessDF <- read.csv("ConcessTable.csv")
CustomerDF <- read.csv("CustomerTable.csv")

# Check dimensions of each file
cat("MainDF:", dim(MainDF), "| StoreDF:", dim(StoreDF),
    "| ConcessDF:", dim(ConcessDF), "| CustomerDF:", dim(CustomerDF), "\n")
```

```{r check_duplicates, echo=FALSE}
# Check for duplicate Cust_IDs in each source table BEFORE merging
cat("Duplicates in MainDF:", sum(duplicated(MainDF$Cust_ID)), "\n")
cat("Duplicates in StoreDF:", sum(duplicated(StoreDF$Cust_ID)), "\n")
cat("Duplicates in ConcessDF:", sum(duplicated(ConcessDF$Cust_ID)), "\n")
cat("Duplicates in CustomerDF:", sum(duplicated(CustomerDF$Cust_ID)), "\n")
```

```{r merge_data, echo=FALSE}
# Create flat file by merging all tables on Cust_ID
FFdf <- MainDF
FFdf <- merge(FFdf, StoreDF, by = "Cust_ID", all.x = TRUE)
FFdf <- merge(FFdf, ConcessDF, by = "Cust_ID", all.x = TRUE)
FFdf <- merge(FFdf, CustomerDF, by = "Cust_ID", all.x = TRUE)

cat("Final flat file dimensions - Rows:", nrow(FFdf), "Columns:", ncol(FFdf), "\n")
cat("Unique Cust_ID:", length(unique(FFdf$Cust_ID)), "| Duplicate rows:", nrow(FFdf) - length(unique(FFdf$Cust_ID)), "\n")
```

## Step 1: Open Data in Your Software of Choice

Create identifier variable, arrange columns, and explore distributions visually.

```{r step1_open_data, echo=FALSE, fig.width=8, fig.height=6}
# Create an identifier variable
FFdf$ID <- 1:nrow(FFdf)

# Reorder columns: Y-variable first, then alphabetically
y_var <- "Y01"
other_vars <- sort(setdiff(names(FFdf), y_var))
FFdf <- FFdf[, c(y_var, other_vars)]

# Histograms for key numeric variables
par(mfrow = c(2, 2))
hist(FFdf$Age, main = "Distribution of Age", xlab = "Age", col = "steelblue", breaks = 20)
hist(FFdf$tenure, main = "Distribution of Tenure", xlab = "Tenure (Years)", col = "steelblue", breaks = 20)
hist(FFdf$NumSeats, main = "Distribution of NumSeats", xlab = "Number of Seats", col = "steelblue")
hist(FFdf$Total.Spent, main = "Distribution of Total Spent", xlab = "Total Spent ($)", col = "steelblue", breaks = 30)
par(mfrow = c(1, 1))

# Response variable distribution
cat("Response Variable (Y01):", table(FFdf$Y01), "\n")
barplot(table(FFdf$Y01), main = "Distribution of Y01 (Response Variable)",
        col = c("coral", "steelblue"), names.arg = c("0 (No)", "1 (Yes)"))
```

**Observations:** Age is roughly normal (30-60), Tenure is right-skewed, NumSeats clusters around 2-4, Total Spent is heavily right-skewed. Response variable Y01 is reasonably balanced.

## Step 2: Review Variables for Common Sense (SME Knowledge)

Standardize variable names and check for unique identifiers.

```{r step2_common_sense, echo=FALSE}
# Clean up variable names - replace dots/spaces with underscores, capitalize
names(FFdf) <- gsub("\\.", "_", names(FFdf))
names(FFdf) <- gsub(" ", "_", names(FFdf))
names(FFdf) <- gsub("(^|_)([a-z])", "\\1\\U\\2", names(FFdf), perl = TRUE)

cat("Dataset:", nrow(FFdf), "rows,", ncol(FFdf), "columns\n")
cat("Unique Cust_ID:", length(unique(FFdf$Cust_ID)), "out of", nrow(FFdf), "rows\n")
```

**Result:** Each row does not necessarily represent a unique customer (175 duplicate Cust_IDs found in MainDF source data).

## Step 3: Review How Software Coded Variables

Convert character variables to factors for proper categorical analysis.

```{r step3_variable_types, echo=FALSE}
# Identify character variables
chr_vars <- names(FFdf)[sapply(FFdf, is.character)]
cat("Character variables to convert:", length(chr_vars), "\n")

# Convert character variables to factors
for(var in chr_vars) {
  FFdf[[var]] <- as.factor(FFdf[[var]])
}

# Show key factor levels
cat("\nSex levels:", levels(FFdf$Sex), "\n")
cat("Marital levels:", levels(FFdf$Marital), "\n")
cat("Account_Type levels:", levels(FFdf$Account_Type), "\n")
```

**Note:** Marital has levels D (Divorced), M (Married), S (Single), U (Unknown). The "U" for Unknown may need special handling.

## Step 4: Data Integrity/Validation Checks

Check for anomalies, bogus values, and data quality issues.

```{r step4_checks, echo=FALSE}
# Check numeric variable ranges
cat("Age range:", range(FFdf$Age, na.rm = TRUE), "\n")
cat("Tenure range:", range(FFdf$Tenure, na.rm = TRUE), "\n")
cat("NumSeats range:", range(FFdf$NumSeats, na.rm = TRUE), "\n")

# Check for 999 placeholder values in DistA
cat("\nDistA values = 999:", sum(FFdf$DistA == 999, na.rm = TRUE), "\n")

# Convert 999 to NA (placeholder for missing)
FFdf$DistA[FFdf$DistA == 999] <- NA
cat("DistA NA count after conversion:", sum(is.na(FFdf$DistA)), "\n")

# Check Survey_Comp for outliers (expected range 0-1)
cat("\nSurvey_Comp range:", range(FFdf$Survey_Comp, na.rm = TRUE), "\n")
cat("Survey_Comp values > 1:", sum(FFdf$Survey_Comp > 1, na.rm = TRUE), "\n")

# Check for redundant columns
cat("\nState_Name unique:", length(unique(FFdf$State_Name)),
    "| State_Loc unique:", length(unique(FFdf$State_Loc)), "\n")
```

**Issues Found:**

- **DistA = 999:** Placeholder values converted to NA
- **Survey_Comp > 1:** Outlier found when expected range is 0-1
- **Age max = 99:** May be placeholder or extreme value - verify with SME
- **Tenure max = 400:** Suspicious value if measured in years - verify units with SME
- **State_Name/State_Loc:** Redundant columns (same info, different format)
- **Marital = "U":** Unknown status - consider treating as NA
- **Cust_ID duplicates:** 175 duplicate IDs found in MainDF source data
- **Address, Name, PhoneNum:** 100% missing - likely removed from CustomerDF for privacy

## Step 5: Handle Dates

Convert Last_Contact datetime and extract useful components.

```{r step5_handle_dates, echo=FALSE}
# Convert Last_Contact to datetime format
FFdf$Last_Contact <- ymd_hms(as.character(FFdf$Last_Contact))

# Extract date components
FFdf$Contact_Year <- year(FFdf$Last_Contact)
FFdf$Contact_Month <- month(FFdf$Last_Contact)
FFdf$Contact_Day <- day(FFdf$Last_Contact)
FFdf$Contact_Weekday <- wday(FFdf$Last_Contact, label = TRUE)
FFdf$Contact_Hour <- hour(FFdf$Last_Contact)

cat("Date components extracted: Contact_Year, Contact_Month, Contact_Day, Contact_Weekday, Contact_Hour\n")
cat("Contact Year range:", range(FFdf$Contact_Year, na.rm = TRUE), "\n")
cat("Contact Hour range:", range(FFdf$Contact_Hour, na.rm = TRUE), "\n")
```

## Summary

```{r summary, echo=FALSE}
cat("Final Dataset:", nrow(FFdf), "rows x", ncol(FFdf), "columns\n")
cat("Total Missing Values:", sum(is.na(FFdf)), "\n")

# Show columns with missing values
na_counts <- colSums(is.na(FFdf))
na_cols <- na_counts[na_counts > 0]
cat("\nColumns with missing values:\n")
print(sort(na_cols, decreasing = TRUE))
```

### Data Quality Issues for Future Steps:

| Issue | Possible action / Action Taken |
|-------|--------------|
| DistA = 999 | Converted to NA |
| Survey_Comp > 1 | Flag for investigation (110 values, max 7.4) |
| Age = 99 | Verify with SME |
| Tenure = 400 | Verify units with SME (400 years unlikely) |
| Marital = "U" | Consider as NA or keep |
| State redundancy | Drop one column |
| ID columns | Exclude from modeling |
| Cust_ID duplicates | 175 duplicates in MainDF - investigate or deduplicate |
| PII columns | Address, Name, PhoneNum 100% missing - exclude |



**Step 6: Handle Categorical Variables - keep as is, combine rare levels, combine similar levels**


```{r step6_categorical, echo=FALSE}
# Get all factor variables
factor_vars <- names(FFdf)[sapply(FFdf, is.factor)]
cat("Factor variables:", length(factor_vars), "\n")

# Function to check for rare levels (less than 5% of data)
check_rare_levels <- function(df, threshold = 0.05) {
  factor_vars <- names(df)[sapply(df, is.factor)]
  rare_info <- list()

  for(var in factor_vars) {
    tbl <- table(df[[var]], useNA = "ifany")
    pct <- prop.table(tbl)
    rare_levels <- names(pct[pct < threshold])
    if(length(rare_levels) > 0) {
      rare_info[[var]] <- data.frame(
        Level = rare_levels,
        Count = as.numeric(tbl[rare_levels]),
        Percent = round(as.numeric(pct[rare_levels]) * 100, 2)
      )
    }
  }
  return(rare_info)
}

# Check for rare levels
rare_levels <- check_rare_levels(FFdf, threshold = 0.05)
cat("\nVariables with rare levels (< 5%):\n")
print(names(rare_levels))

# Display level distributions for key categorical variables
cat("\n--- Sex Distribution ---\n")
print(table(FFdf$Sex, useNA = "ifany"))

cat("\n--- Marital Distribution ---\n")
print(table(FFdf$Marital, useNA = "ifany"))

cat("\n--- Account_Type Distribution ---\n")
print(table(FFdf$Account_Type, useNA = "ifany"))

cat("\n--- Educational_Level Distribution ---\n")
print(table(FFdf$Educational_Level, useNA = "ifany"))
```

```{r step6_handle_categorical, echo=FALSE}
# Handle Marital "U" (Unknown) - treat as NA for analysis purposes
# Create backup first
FFdf$Marital_Original <- FFdf$Marital

# Option: Convert "U" to NA (uncomment if desired)
# FFdf$Marital[FFdf$Marital == "U"] <- NA
# FFdf$Marital <- droplevels(FFdf$Marital)

cat("Marital 'U' (Unknown) count:", sum(FFdf$Marital == "U", na.rm = TRUE), "\n")
cat("Decision: Keep 'U' as separate level for now - may represent meaningful unknown status\n")

# Check for levels that might mean the same thing
cat("\n--- State_Name levels ---\n")
print(length(levels(FFdf$State_Name)))
cat("Number of unique states:", length(unique(FFdf$State_Name)), "\n")
```

```{r step6_lump_rare_levels, echo=FALSE}
# Fix Mult_Loc "Y" -> "Yes" BEFORE lumping, so both merge into "Yes" (not separately into "Other")
levels(FFdf$Mult_Loc)[levels(FFdf$Mult_Loc) == "Y"] <- "Yes"

# Lump rare factor levels into "Other" using forcats::fct_lump_prop()
# Threshold: levels appearing in < 5% of non-NA observations are combined

# Store original levels for reference
factor_vars_to_lump <- names(FFdf)[sapply(FFdf, is.factor)]

# Exclude variables we handle separately (State_Name -> Region, Marital already reviewed)
skip_vars <- c("State_Name", "State_Loc", "Marital", "Marital_Original","Educational_Level","Seating_Location","Favorite_Caps_Player","Job_Sector")
lump_vars <- setdiff(factor_vars_to_lump, skip_vars)

cat("=== LUMPING RARE LEVELS (< 5%) INTO 'Other' ===\n\n")

for(var in lump_vars) {
  original_levels <- nlevels(FFdf[[var]])
  FFdf[[var]] <- fct_lump_prop(FFdf[[var]], prop = 0.05, other_level = "Other")
  new_levels <- nlevels(FFdf[[var]])

  if(original_levels != new_levels) {
    cat(var, ": ", original_levels, " -> ", new_levels, " levels\n", sep = "")
    print(table(FFdf[[var]], useNA = "ifany"))
    cat("\n")
  }
}

cat("Lumping complete.\n")
```

```{r step6_group_states_by_region, echo=FALSE}


# Define region mappings
northeast <- c("Connecticut", "Maine", "Massachusetts", "New Hampshire",
               "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")

midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin",
             "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska",
             "North Dakota", "South Dakota")

south <- c("Delaware", "Florida", "Georgia", "North Carolina",
           "South Carolina", "West Virginia",
           "Alabama", "Kentucky", "Mississippi", "Tennessee",
           "Arkansas", "Louisiana", "Oklahoma", "Texas")

west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico",
          "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington")

dcarea <- c("Maryland","Virginia", "District of Columbia")

# Create Region variable
FFdf$Region <- case_when(
  FFdf$State_Name %in% northeast ~ "Northeast",
  FFdf$State_Name %in% midwest ~ "Midwest",
  FFdf$State_Name %in% south ~ "South",
  FFdf$State_Name %in% west ~ "West",
  FFdf$State_Name %in% dcarea ~ "DCArea",
  TRUE ~ "Other"  # Catch any unmatched states
)

# Convert to factor
FFdf$Region <- as.factor(FFdf$Region)

# Display region distribution
cat("--- Region Distribution (grouped from State_Name) ---\n")
print(table(FFdf$Region, useNA = "ifany"))
cat("\nRegion percentages:\n")
print(round(prop.table(table(FFdf$Region)) * 100, 2))

# Check if any states weren't mapped
unmatched_states <- unique(FFdf$State_Name[FFdf$Region == "Other"])
if(length(unmatched_states) > 0) {
  cat("\nWarning - Unmatched states assigned to 'Other':\n")
  print(unmatched_states)
} else {
  cat("\nAll states successfully mapped to regions.\n")
}
```

**Step 6 Observations:**

- Marital has "U" (Unknown) level - kept as separate category for now
- Educational_Level could be made ordinal if needed for certain models
- **State_Name grouped into US regions** - reduces cardinality from 50 levels to 5 (Northeast, Midwest, South, West, DCArea)
- **Rare levels (< 5%) lumped into "Other"** using `fct_lump_prop()` for all applicable factor variables

---

**Step 7: Remove Zero-Variance Predictors**



```{r step7_zero_variance, echo=FALSE}
# Function to identify zero-variance columns
find_zero_variance <- function(df) {
  zv_cols <- c()
  for(col in names(df)) {
    unique_vals <- length(unique(na.omit(df[[col]])))
    if(unique_vals <= 1) {
      zv_cols <- c(zv_cols, col)
    }
  }
  return(zv_cols)
}

# Find zero-variance columns
zero_var_cols <- find_zero_variance(FFdf)
cat("Zero-variance columns found:", length(zero_var_cols), "\n")

if(length(zero_var_cols) > 0) {
  cat("Columns with zero variance:\n")
  print(zero_var_cols)

  # Store in excluded columns list
  excluded_cols <- zero_var_cols

  # Remove zero-variance columns
  FFdf <- FFdf[, !names(FFdf) %in% zero_var_cols]
  cat("\nRemoved", length(zero_var_cols), "zero-variance columns\n")
} else {
  cat("No zero-variance columns found\n")
  excluded_cols <- c()
}

cat("Remaining columns:", ncol(FFdf), "\n")
```

**Step 7 Results:**

The following 8 zero-variance columns were identified and removed:

- **Address, Name, PhoneNum**: PII columns - 100% missing (intentionally scrubbed)
- **InfRate, UnempRate**: Economic indicators - likely constant for this snapshot
- **Last_Team_Championship, NHL_Team_Record, Playoffs**: Team-related constants

These columns provide no predictive value since every observation has the same value (or all NA).

---

## Class 4: Data Cleaning Process: Steps 8-11

**Step 8: Handle Near Zero-Variance Predictors**


```{r step8_near_zero_variance, echo=FALSE}
# Function to find near-zero variance columns
# A column is NZV if one value dominates (e.g., >95% of values)
find_near_zero_variance <- function(df, threshold = 0.95) {
  nzv_info <- data.frame(
    Variable = character(),
    DominantValue = character(),
    DominantPct = numeric(),
    UniqueValues = integer(),
    stringsAsFactors = FALSE
  )

  for(col in names(df)) {
    if(is.numeric(df[[col]]) || is.factor(df[[col]])) {
      tbl <- table(df[[col]], useNA = "no")
      if(length(tbl) > 0) {
        max_pct <- max(tbl) / sum(tbl)
        if(max_pct >= threshold) {
          dominant_val <- names(tbl)[which.max(tbl)]
          nzv_info <- rbind(nzv_info, data.frame(
            Variable = col,
            DominantValue = as.character(dominant_val),
            DominantPct = round(max_pct * 100, 2),
            UniqueValues = length(tbl),
            stringsAsFactors = FALSE
          ))
        }
      }
    }
  }
  return(nzv_info)
}

# Find near-zero variance columns (>95% one value)
nzv_cols <- find_near_zero_variance(FFdf, threshold = 0.95)
cat("Near-zero variance columns (>95% one value):\n")
print(nzv_cols)
```

```{r step8_handle_nzv, echo=FALSE}
# Decision: Flag NZV columns but don't remove yet
# These may still be useful predictors depending on modeling goals
nzv_variables <- nzv_cols$Variable

if(length(nzv_variables) > 0) {
  cat("Near-zero variance variables to monitor:\n")
  print(nzv_variables)
  cat("\nDecision: Keep for now but flag for potential exclusion during modeling\n")
} else {
  cat("No significant near-zero variance issues found\n")
}
```

**Step 8 Results:**

Near-zero variance columns identified (>95% one value):

| Variable | Dominant Value | Dominant % |
|----------|----------------|------------|
| Additional_Seats | 0 | 96.99% |
| Mult_Loc | No | 96.99% |

**Observations:**

- **Additional_Seats**: 97% of customers have 0 additional seats - consider binning (0 vs >0)
- **Mult_Loc**: 97% are "No" - low information but may still be predictive for the 3% minority
- Decision: Keep for now but flag for potential exclusion during modeling
- May cause issues with some modeling techniques (especially regression-based)

---

**Step 9: Remove Redundant Columns and Linear Combination Columns**


```{r step9_redundant, echo=FALSE}
# Check for redundant categorical columns (State_Name vs State_Loc)
cat("--- Checking State_Name vs State_Loc redundancy ---\n")
if("State_Name" %in% names(FFdf) && "State_Loc" %in% names(FFdf)) {
  # Check if they're perfectly correlated
  state_comparison <- table(FFdf$State_Name, FFdf$State_Loc)
  cat("State_Name unique values:", length(unique(FFdf$State_Name)), "\n")
  cat("State_Loc unique values:", length(unique(FFdf$State_Loc)), "\n")

  # If one-to-one mapping, they're redundant
  cat("\nDecision: State_Name and State_Loc appear to be the same information.\n")
  cat("Removing State_Loc (keeping State_Name)\n")

  excluded_cols <- c(excluded_cols, "State_Loc")
  FFdf$State_Loc <- NULL
}

# Check correlation matrix for numeric variables
numeric_vars <- names(FFdf)[sapply(FFdf, is.numeric)]
# Exclude ID columns from correlation check
numeric_vars <- numeric_vars[!numeric_vars %in% c("ID", "Cust_ID")]

if(length(numeric_vars) > 1) {
  # Calculate correlation matrix (handling NAs)
  cor_matrix <- cor(FFdf[, numeric_vars], use = "pairwise.complete.obs")

  # Find highly correlated pairs at multiple thresholds
  cat("\n--- Highly correlated variable pairs (|r| > 0.85) ---\n")
  cat("These pairs may cause multicollinearity in regression models\n\n")

  high_cor_85 <- which(abs(cor_matrix) > 0.85 & abs(cor_matrix) < 1, arr.ind = TRUE)

  if(nrow(high_cor_85) > 0) {
    cor_pairs <- data.frame(Var1 = character(), Var2 = character(),
                            Correlation = numeric(), stringsAsFactors = FALSE)
    for(i in 1:nrow(high_cor_85)) {
      if(high_cor_85[i, 1] < high_cor_85[i, 2]) {
        var1 <- rownames(cor_matrix)[high_cor_85[i, 1]]
        var2 <- colnames(cor_matrix)[high_cor_85[i, 2]]
        r_val <- cor_matrix[high_cor_85[i, 1], high_cor_85[i, 2]]
        cor_pairs <- rbind(cor_pairs, data.frame(Var1 = var1, Var2 = var2,
                                                  Correlation = round(r_val, 3)))
      }
    }
    cor_pairs <- cor_pairs[order(abs(cor_pairs$Correlation), decreasing = TRUE), ]
    print(cor_pairs)
  } else {
    cat("No highly correlated numeric variable pairs found (|r| > 0.85)\n")
  }
}
```

```{r step9_correlation_plot, echo=FALSE, fig.width=10, fig.height=8}
# Visualize correlation matrix for numeric variables
if(length(numeric_vars) > 2) {
  # Subset to variables with fewer missing values for cleaner plot
  complete_vars <- numeric_vars[colSums(is.na(FFdf[, numeric_vars])) < nrow(FFdf) * 0.5]

  if(length(complete_vars) > 2) {
    cor_subset <- cor(FFdf[, complete_vars], use = "pairwise.complete.obs")
    corrplot(cor_subset, method = "color", type = "upper",
             tl.cex = 0.7, tl.col = "black",
             title = "Correlation Matrix - Numeric Variables",
             mar = c(0, 0, 2, 0))
  }
}
```

```{r step9_handle_multicollinearity, echo=FALSE}
# Address multicollinearity based on correlation matrix analysis
cat("=== MULTICOLLINEARITY ANALYSIS ===\n\n")

# Identify correlated variable clusters from the correlation matrix
cat("Cluster 1: Spending & Visit variables\n")
cat("  - Rep_Visits <-> Total_Spent: r = 0.947 (very strong positive)\n")
cat("  - Team_Store_Total <-> Total_Spent: r = 0.853 (strong positive)\n")
cat("  Recommendation: Consider removing Rep_Visits or Total_Spent\n\n")

cat("Cluster 2: Concession & Seating\n")
cat("  - Concession_Total <-> NumSeats: r = 0.915 (strong positive)\n")
cat("  Recommendation: Makes business sense - more seats = more concessions\n\n")

cat("Cluster 3: Attendance pairs\n")
cat("  - Weekday_Attended <-> Weekday_Sold: r = -0.946 (strong NEGATIVE)\n")
cat("  Note: Negative correlation suggests inverse relationship\n")
cat("  Recommendation: Keep both - they capture different behaviors\n\n")

# Create list of variables to flag for multicollinearity
multicollinear_vars <- c("Rep_Visits", "Team_Store_Total")
cat("Variables flagged for potential removal due to multicollinearity:\n")
print(multicollinear_vars)

# Option: Remove highly correlated variables (uncomment to execute)
# FFdf <- FFdf[, !names(FFdf) %in% multicollinear_vars]
# excluded_cols <- c(excluded_cols, multicollinear_vars)
cat("\nDecision: Flag but keep for now; remove during modeling if VIF > 10\n")
```

**Step 9 Observations:**

Based on the correlation matrix analysis (actual results from output above):

| Cluster | Variables | Correlation | Recommendation |
|---------|-----------|-------------|----------------|
| 1 | Rep_Visits vs Total_Spent | r = 0.947 | Remove Rep_Visits |
| 1 | Team_Store_Total vs Total_Spent | r = 0.853 | Monitor for VIF |
| 2 | Concession_Total vs NumSeats | r = 0.915 | Keep - business logic |
| 3 | Weekday_Attended vs Weekday_Sold | r = -0.946 | Keep both - inverse relationship |
| - | State_Loc vs State_Name | Redundant | **REMOVED** |

**Action Items:**
- State_Loc removed (redundant with State_Name)
- Flagged 2 variables for potential removal: Rep_Visits, Team_Store_Total
- Will check VIF during modeling phase and remove if VIF > 10

---

**Step 10: Search for Outliers and Initial Search for Missing Values**

```{r step10_outliers, echo=FALSE, fig.width=10, fig.height=8}
# Boxplots for key numeric variables to identify outliers
par(mfrow = c(2, 3))

# Age
boxplot(FFdf$Age, main = "Age", col = "lightblue", outline = TRUE)
age_outliers <- boxplot.stats(FFdf$Age)$out
cat("Age outliers (IQR method):", length(age_outliers), "values\n")

# Tenure
boxplot(FFdf$Tenure, main = "Tenure", col = "lightblue", outline = TRUE)
tenure_outliers <- boxplot.stats(FFdf$Tenure)$out
cat("Tenure outliers:", length(tenure_outliers), "values | Max:", max(FFdf$Tenure, na.rm = TRUE), "\n")

# Total_Spent
boxplot(FFdf$Total_Spent, main = "Total_Spent", col = "lightblue", outline = TRUE)

# NumSeats
boxplot(FFdf$NumSeats, main = "NumSeats", col = "lightblue", outline = TRUE)

# Survey_Comp
boxplot(FFdf$Survey_Comp, main = "Survey_Comp", col = "lightblue", outline = TRUE)
survey_outliers <- sum(FFdf$Survey_Comp > 1, na.rm = TRUE)
cat("Survey_Comp values > 1:", survey_outliers, "\n")

# DistA
boxplot(FFdf$DistA, main = "DistA", col = "lightblue", outline = TRUE)

par(mfrow = c(1, 1))
```

```{r step10_missing_assessment, echo=FALSE, fig.width=10, fig.height=6}
# Comprehensive missing data assessment
cat("=== MISSING DATA ASSESSMENT ===\n\n")

# Total missing values
total_missing <- sum(is.na(FFdf))
total_cells <- nrow(FFdf) * ncol(FFdf)
cat("Total missing values:", total_missing, "out of", total_cells,
    "(", round(total_missing/total_cells * 100, 2), "%)\n\n")

# Missing by column
na_by_col <- colSums(is.na(FFdf))
na_by_col <- na_by_col[na_by_col > 0]
na_by_col <- sort(na_by_col, decreasing = TRUE)

cat("Columns with missing values:\n")
na_summary <- data.frame(
  Variable = names(na_by_col),
  Missing_Count = as.numeric(na_by_col),
  Missing_Pct = round(as.numeric(na_by_col) / nrow(FFdf) * 100, 2)
)
print(na_summary)

```

```{r step10_outlier_decisions, echo=FALSE}
# Document outlier decisions
cat("\n=== OUTLIER DECISIONS ===\n\n")

# Tenure = 400 (suspicious if years)
cat("1. Tenure max = 400:\n")
cat("   - If measured in years, this is impossible\n")
cat("   - May be measured in months (400 months = 33 years - plausible)\n")
cat("   - ACTION: Verify units with SME; flag for review\n\n")

# Age = 99
cat("2. Age = 99:\n")
cat("   - Could be real (elderly customer) or placeholder\n")
cat("   - ACTION: Verify with SME; consider if 99 is data entry default\n\n")

# Survey_Comp > 1
cat("3. Survey_Comp values > 1 (expected 0-1 range):\n")
cat("   - Count:", sum(FFdf$Survey_Comp > 1, na.rm = TRUE), "\n")
cat("   - Max value:", max(FFdf$Survey_Comp, na.rm = TRUE), "\n")
cat("   - ACTION: Possible scale issue; cap at 1 or investigate data source\n\n")

# Create outlier flags for further analysis
FFdf$Flag_Tenure_High <- ifelse(FFdf$Tenure > 100, 1, 0)
FFdf$Flag_Survey_Invalid <- ifelse(FFdf$Survey_Comp > 1, 1, 0)

cat("Created outlier flag variables: Flag_Tenure_High, Flag_Survey_Invalid\n")
```

---

**Step 11: Sanity Check Using Decision Tree (1 to 2 splits)**


```{r step11_decision_tree, echo=FALSE, fig.width=10, fig.height=8}
# Prepare data for decision tree
# Exclude ID columns, flag variables, and high-cardinality variables
exclude_from_tree <- c("ID", "Cust_ID", "Flag_Tenure_High", "Flag_Survey_Invalid",
                       "Marital_Original", "Last_Contact", "Contact_Year",
                       "Contact_Month", "Contact_Day", "Contact_Weekday", "Contact_Hour",
                       "State_Name", "Zip_Codes")

tree_vars <- names(FFdf)[!names(FFdf) %in% exclude_from_tree]
tree_data <- FFdf[, tree_vars]

# Remove rows with NA in response variable
tree_data <- tree_data[!is.na(tree_data$Y01), ]

# Convert Y01 to factor for classification tree
tree_data$Y01 <- as.factor(tree_data$Y01)

# Build a simple decision tree (max 2-3 splits for sanity check)
# Uses Region variable (from Step 6) instead of high-cardinality State_Name/Zip_Codes
sanity_tree <- rpart(Y01 ~ .,
                     data = tree_data,
                     method = "class",
                     control = rpart.control(maxdepth = 3, minsplit = 20, cp = 0.01))

# Plot the tree
rpart.plot(sanity_tree,
           main = "Sanity Check Decision Tree (max depth = 3)",
           extra = 104,  # Show percentage and count
           box.palette = "RdYlGn")

# Variable importance
cat("\n=== VARIABLE IMPORTANCE ===\n")
if(length(sanity_tree$variable.importance) > 0) {
  var_imp <- sort(sanity_tree$variable.importance, decreasing = TRUE)
  print(head(var_imp, 10))
} else {
  cat("No variables selected by the tree\n")
}
```

```{r step11_tree_analysis, echo=FALSE}
# Analyze tree results
cat("\n=== SANITY CHECK ANALYSIS ===\n\n")

# Check tree performance
predictions <- predict(sanity_tree, tree_data, type = "class")
accuracy <- mean(predictions == tree_data$Y01, na.rm = TRUE)
cat("Tree accuracy:", round(accuracy * 100, 2), "%\n")

if(accuracy > 0.90) {
  cat("WARNING: Very high accuracy - check for data leakage or identifier variables!\n")
} else {
  cat("Accuracy is reasonable - no obvious data leakage detected\n")
}
```

**Step 11 Results:**

The decision tree uses Region (from Step 6) instead of raw State_Name/Zip_Codes to avoid overfitting from high-cardinality variables.

Top Variable Importance:

| Variable | Importance |
|----------|------------|
| Region | 1505.2 |
| Favorite_Team | 253.5 |
| PerUsed | 245.2 |
| DistA | 196.8 |
| Tenure | 187.7 |

**Analysis:**

- **Accuracy ~88%** - reasonable, no obvious data leakage
- **Region is the dominant predictor** - geographic location strongly predicts Y01
- **No high-cardinality variables** causing artificial inflation of accuracy

---

## Summary of Steps 6-11 (Data Cleaning Complete)

```{r final_summary, echo=FALSE}
cat("=== DATA CLEANING SUMMARY (Steps 6-11) ===\n\n")

cat("Step 6 - Handle Categorical Variables:\n")
cat("  - Rare factor levels (< 5%) lumped into 'Other' via fct_lump_prop()\n")
cat("  - Marital 'U' kept as separate category\n")
cat("  - State_Name grouped into US Census regions\n\n")

cat("Step 7 - Zero-Variance Predictors:\n")
cat("  - Columns removed:", length(zero_var_cols), "\n\n")

cat("Step 8 - Near Zero-Variance Predictors:\n")
cat("  - Variables flagged:", length(nzv_variables), "\n")
cat("  - Decision: Keep for now but monitor during modeling\n\n")

cat("Step 9 - Redundant Columns:\n")
cat("  - State_Loc removed (redundant with State_Name)\n")
cat("  - Correlation matrix reviewed for multicollinearity\n\n")

cat("Step 10 - Outliers & Missing Data:\n")
cat("  - Total missing values:", sum(is.na(FFdf)), "\n")
cat("  - Outlier flags created for Tenure and Survey_Comp\n")
cat("  - Missing data summary table generated\n\n")

cat("Step 11 - Decision Tree Sanity Check:\n")
cat("  - Tree accuracy:", round(accuracy * 100, 2), "%\n")
cat("  - Review variable importance for potential data leakage\n\n")

cat("Final dataset dimensions:", nrow(FFdf), "rows x", ncol(FFdf), "columns\n")

# Save cleaned dataset for next phase (Missing Data)
tryCatch({
  write.csv(FFdf, "FFdf_cleaned.csv", row.names = FALSE)
  cat("\nCleaned dataset saved to: FFdf_cleaned.csv\n")
}, error = function(e) cat("\nNote: Could not save FFdf_cleaned.csv (file may be open):", conditionMessage(e), "\n"))
```

### Issues Identified for Further Action:

| Step | Issue | Recommendation |
|------|-------|----------------|
| 6 | Marital "U" unknown | Keep as category or convert to NA during imputation |
| 6 | Rare factor levels (< 5%) | **RESOLVED:** Lumped into "Other" via fct_lump_prop() |
| 6 | State_Name high cardinality | **RESOLVED:** Grouped into US Census regions |
| 7 | Zero-variance columns | Removed from dataset |
| 8 | Near-zero variance | Monitor during modeling; consider binning |
| 9 | State_Loc redundant | Removed |
| 10 | Tenure = 400 | Verify units with SME (years vs months?) |
| 10 | Survey_Comp > 1 | Investigate scale/cap values at 1 |
| 10 | Missing data patterns | Address in Missing Data phase (Class 5+) |
| 11 | Tree predictors | **RESOLVED:** Using Region variable instead of State_Name |

---

## Missing Data Analysis: Steps 1-6


### MD Step 1: Identify Missing Data

Rename to MDdf and count all missing values by variable.

```{r md_step1_identify, echo=TRUE, fig.width=10, fig.height=8}
MDdf <- FFdf

missing_summary <- data.frame(
  Variable     = names(MDdf),
  Missing_Count = colSums(is.na(MDdf)),
  Missing_Pct  = round(colSums(is.na(MDdf)) / nrow(MDdf) * 100, 1)
) %>% filter(Missing_Count > 0) %>% arrange(desc(Missing_Count))

cat("Total missing:", sum(is.na(MDdf)), "of", nrow(MDdf)*ncol(MDdf), "cells (",
    round(sum(is.na(MDdf))/(nrow(MDdf)*ncol(MDdf))*100,1), "%)\n")

flextable(missing_summary) %>%
  set_header_labels(Variable="Variable", Missing_Count="Missing (n)", Missing_Pct="Missing (%)") %>%
  autofit()
```

**Observations:**

- 12.2% overall missing; 7 CustomerDF variables (Educational_Level, Favorite_Caps_Player, Favorite_Sport, Job_Sector, Mode_Of_Transport, Team_B_STH, Team_C_STH) all missing at exactly 70% — likely a block from customers who skipped a supplemental survey.
- Net_Worth_True (61%) and HouseHold_Income_True (60%) are high — sensitive financial fields.
- DistA (30.6%) includes 999-placeholder NAs converted in Step 4.
- Moderate missingness (5–15%): Marital, Age, Rep_Name, Sex, Tenure, Rep_Visits, Rep_Calls, Num_Children.

---

### MD Step 2: Mark Missing Data

Create binary indicator variables (M_VarName: 1 = missing, 0 = present) for all variables with NAs. Indicators allow MCAR/MAR testing and preserve pre-imputation missingness structure.

```{r md_step2_mark, echo=TRUE, fig.width=10, fig.height=8}
vars_with_na <- names(MDdf)[colSums(is.na(MDdf)) > 0]
vars_with_na <- vars_with_na[!vars_with_na %in% c("Marital_Original", "Flag_Tenure_High")]

for(var in vars_with_na) {
  MDdf[[paste0("M_", var)]] <- ifelse(is.na(MDdf[[var]]), 1, 0)
}

indicator_vars <- grep("^M_", names(MDdf), value = TRUE)
cat("Indicator variables created:", length(indicator_vars), "\n")

# Correlation of indicators — high r = missing as a block
indicator_matrix <- MDdf[, indicator_vars]
indicator_matrix <- indicator_matrix[, sapply(indicator_matrix, stats::var) > 0]
cor_indicators <- cor(indicator_matrix)

corrplot(cor_indicators, method = "color", type = "upper",
         tl.cex = 0.7, tl.col = "black",
         title = "Correlation of Missing Data Indicators",
         mar = c(0, 0, 2, 0))
```

**Observations:**

- The 7 CustomerDF block variables have perfect indicator correlation (r = 1.0), confirming they are missing together.
- Net_Worth_True and HouseHold_Income_True indicators are moderately correlated — missing together as financial fields.

---

### MD Step 3: Clean Up Obvious Mistakes

Fix values that are wrong or that should be coded as NA.

```{r md_step3_cleanup, echo=TRUE}
# Survey_Comp > 1: cap at 1.0 (proportion cannot exceed 100%)
cat("Survey_Comp > 1 count:", sum(MDdf$Survey_Comp > 1, na.rm = TRUE), "| max:", max(MDdf$Survey_Comp, na.rm = TRUE), "\n")
MDdf$Survey_Comp[MDdf$Survey_Comp > 1] <- 1.0
cat("Survey_Comp after cap:", range(MDdf$Survey_Comp, na.rm = TRUE), "\n\n")

# Marital "U": keep as valid level — unknown status is informative
cat("Marital 'U' count:", sum(MDdf$Marital == "U", na.rm = TRUE), "— KEPT as distinct level\n")
cat("Marital true NAs:", sum(is.na(MDdf$Marital)), "\n\n")

# Tenure / Age extremes: flag only, leave for SME
cat("Tenure > 100:", sum(MDdf$Tenure > 100, na.rm = TRUE), "| Age == 99:", sum(MDdf$Age == 99, na.rm = TRUE), "— flagged, left as-is\n")
cat("Total missing after cleanup:", sum(is.na(MDdf)), "\n")
```

**Observations:**

- Survey_Comp: 110 values capped at 1.0 (data entry errors on a 0–1 proportion scale).
- Marital "U" (248 values): kept — preserves information, lets model treat "Unknown" as its own category. True NAs still go to MICE.
- Tenure = 400 / Age = 99: left pending SME confirmation of units.

---

### MD Step 4: Make Easy Decisions on Rows/Columns

Define which columns are excluded from imputation (non-analytical only). No rows excluded. All variables with missing data — including the 70% block — are imputed.

```{r md_step4_decisions, echo=TRUE}
id_cols       <- c("ID", "Cust_ID")
backup_cols   <- c("Marital_Original", "Flag_Tenure_High", "Flag_Survey_Invalid")
date_cols     <- c("Last_Contact","Contact_Year","Contact_Month","Contact_Day","Contact_Weekday","Contact_Hour")
high_card_cols <- c("State_Name", "Zip_Codes", "Seating_Location")
indicator_cols <- grep("^M_", names(MDdf), value = TRUE)

all_exclude <- c(id_cols, backup_cols, date_cols, high_card_cols, indicator_cols)
impute_vars  <- names(MDdf)[!names(MDdf) %in% all_exclude]
impute_with_na <- impute_vars[colSums(is.na(MDdf[, impute_vars])) > 0]

cat("Columns excluded from imputation:", length(all_exclude), "\n")
cat("Imputation-eligible variables:", length(impute_vars), "| with NAs:", length(impute_with_na), "\n\n")

# Row missing summary
row_na <- rowSums(is.na(MDdf[, impute_vars]))
cat("Row NA distribution among eligible columns:\n")
cat("  0 missing:", sum(row_na == 0), "| 1-3:", sum(row_na >= 1 & row_na <= 3),
    "| 4-6:", sum(row_na >= 4 & row_na <= 6), "| 7+:", sum(row_na > 6), "\n")
cat("Decision: No rows excluded.\n")
```

**Observations:**

- Excluded: ID/Cust_ID, derived backup/flag columns, date components, high-cardinality cols (State_Name, Zip_Codes, Seating_Location), all M_ indicators.
- All variables with missing data are kept for imputation — including the 70% block — to observe the full effect on modeling.

---

### MD Step 5: Assess Missingness Patterns (MCAR vs MAR)

Use decision trees predicting each variable's M_ indicator. If the tree finds splits, missingness is predictable from other data → MAR. No splits → likely MCAR.

```{r md_step5_trees, echo=TRUE, fig.width=10, fig.height=7}
predictor_candidates <- impute_vars[colSums(is.na(MDdf[, impute_vars])) == 0]
complete_predictors  <- MDdf[, predictor_candidates]

# Drop high-cardinality factors and logical columns
complete_predictors <- complete_predictors[, sapply(complete_predictors, function(x)
  !(is.factor(x) && nlevels(x) > 30))]
complete_predictors <- as.data.frame(lapply(complete_predictors, function(x)
  if(is.logical(x)) as.integer(x) else x))

results <- data.frame(Variable=character(), Splits=integer(), Accuracy=numeric(),
                      Top_Predictor=character(), Assessment=character(), stringsAsFactors=FALSE)

for(var in impute_with_na) {
  target <- factor(MDdf[[paste0("M_", var)]], levels=c(0,1), labels=c("Present","Missing"))
  tree_data <- cbind(Target=target, complete_predictors)
  tree_model <- rpart(Target ~ ., data=tree_data, method="class",
                      control=rpart.control(maxdepth=3, minsplit=50, cp=0.01))

  n_splits <- nrow(tree_model$frame[tree_model$frame$var != "<leaf>", ])
  acc      <- round(mean(predict(tree_model, tree_data, type="class") == target) * 100, 1)
  top_pred <- if(length(tree_model$variable.importance) > 0) names(tree_model$variable.importance)[1] else "None"
  assessment <- if(n_splits == 0) "Likely MCAR" else if(acc > 85) "MAR" else "Possibly MAR"

  results <- rbind(results, data.frame(Variable=var, Splits=n_splits, Accuracy=acc,
                                       Top_Predictor=top_pred, Assessment=assessment,
                                       stringsAsFactors=FALSE))
}

flextable(results) %>%
  set_header_labels(Variable="Variable", Splits="Splits", Accuracy="Accuracy (%)",
                    Top_Predictor="Top Predictor", Assessment="Assessment") %>%
  color(~ Assessment == "MAR",         ~ Assessment, color="red") %>%
  color(~ Assessment == "Likely MCAR", ~ Assessment, color="darkgreen") %>%
  color(~ Assessment == "Possibly MAR",~ Assessment, color="orange") %>%
  autofit()
```

```{r md_step5_tree_plots, echo=TRUE, fig.width=10, fig.height=7}
vars_with_splits <- results$Variable[results$Splits > 0]
for(var in vars_with_splits[1:min(4, length(vars_with_splits))]) {
  target     <- factor(MDdf[[paste0("M_", var)]], levels=c(0,1), labels=c("Present","Missing"))
  tree_model <- rpart(Target ~ ., data=cbind(Target=target, complete_predictors), method="class",
                      control=rpart.control(maxdepth=3, minsplit=50, cp=0.01))
  rpart.plot(tree_model, main=paste("Missingness:", var, "—", results$Assessment[results$Variable==var]),
             extra=104, box.palette="RdYlGn")
}
```

**Observations:**

- **MAR** (splits found): DistA, Most_Purch_Concession — missingness predicted by other variables.
- **Likely MCAR** (no splits): variables where missingness is unpredictable from observed data.
- **7 CustomerDF block variables**: likely MNAR (customers who skipped the survey differ systematically). Imputed anyway to observe effect; interpret with caution.

---

### MD Step 6: Simple (Univariate) Imputation

Handle variables with small missingness using simple methods before MICE. **Management dictates** stochastic (independent) imputation for Rep_Calls and Rep_Name, and regression-based imputation for Age.

| Variable | % Missing | Method |
|----------|-----------|--------|
| Num_Children | 4.3% | Median |
| Rep_Visits | 5.4% | Median |
| Most_Purch_Concession | 0.9% | Mode |
| Sex | 7.1% | Mode |
| Rep_Calls | 4.6% | **Mgmt Dictate** — Stochastic (independent) |
| Rep_Name | 9.2% | **Mgmt Dictate** — Stochastic (independent) |
| Age | 10.4% | **Mgmt Dictate** — Regression (other vars) |
| Remaining 12 vars | 6–70% | → MICE (Step 7) |

```{r md_step6_simple, echo=TRUE}
get_mode <- function(x) { x <- x[!is.na(x)]; ux <- unique(x); ux[which.max(tabulate(match(x, ux)))] }

# --- Median imputation ---
median_nc <- median(MDdf$Num_Children, na.rm=TRUE)
MDdf$Num_Children[is.na(MDdf$Num_Children)] <- median_nc
cat("Num_Children: imputed with median =", median_nc, "| NAs remaining:", sum(is.na(MDdf$Num_Children)), "\n")

median_rv <- median(MDdf$Rep_Visits, na.rm=TRUE)
MDdf$Rep_Visits[is.na(MDdf$Rep_Visits)] <- median_rv
cat("Rep_Visits: imputed with median =", round(median_rv,2), "| NAs remaining:", sum(is.na(MDdf$Rep_Visits)), "\n\n")

# --- Mode imputation ---
mode_mpc <- get_mode(MDdf$Most_Purch_Concession)
MDdf$Most_Purch_Concession[is.na(MDdf$Most_Purch_Concession)] <- mode_mpc
cat("Most_Purch_Concession: imputed with mode =", as.character(mode_mpc), "| NAs:", sum(is.na(MDdf$Most_Purch_Concession)), "\n")

mode_sex <- get_mode(MDdf$Sex)
MDdf$Sex[is.na(MDdf$Sex)] <- mode_sex
cat("Sex: imputed with mode =", as.character(mode_sex), "| NAs:", sum(is.na(MDdf$Sex)), "\n\n")

# --- Mgmt Dictate: Stochastic (independent, sample with replacement) ---
set.seed(42)
observed_rc <- MDdf$Rep_Calls[!is.na(MDdf$Rep_Calls)]
MDdf$Rep_Calls[is.na(MDdf$Rep_Calls)] <- sample(observed_rc, sum(is.na(MDdf$Rep_Calls)), replace=TRUE)
cat("Rep_Calls: stochastic imputation | NAs:", sum(is.na(MDdf$Rep_Calls)), "\n")

set.seed(123)
observed_rn    <- MDdf$Rep_Name[!is.na(MDdf$Rep_Name)]
level_props    <- prop.table(table(observed_rn))
MDdf$Rep_Name[is.na(MDdf$Rep_Name)] <- sample(names(level_props), sum(is.na(MDdf$Rep_Name)),
                                               replace=TRUE, prob=as.numeric(level_props))
cat("Rep_Name: stochastic imputation (proportional) | NAs:", sum(is.na(MDdf$Rep_Name)), "\n")
```

```{r md_step6_age, echo=TRUE}
# --- Mgmt Dictate: Age via linear regression on other variables ---
age_predictors  <- c("Y01","Total_Spent","NumSeats","Tenure","First_Year_STH",
                     "Num_Children","Rep_Calls","Rep_Visits","Concession_Total","Survey_Comp")
available_preds <- age_predictors[sapply(age_predictors, function(v) v %in% names(MDdf) && sum(is.na(MDdf[[v]]))==0)]

train_idx  <- !is.na(MDdf$Age)
age_model  <- lm(Age ~ ., data=MDdf[train_idx, c("Age", available_preds)])
pred_age   <- predict(age_model, newdata=MDdf[!train_idx, available_preds])
pred_age   <- pmin(pmax(round(pred_age), min(MDdf$Age, na.rm=TRUE)), max(MDdf$Age, na.rm=TRUE))

MDdf$Age[!train_idx] <- pred_age
cat("Age: regression imputation (", length(available_preds), "predictors) | NAs:", sum(is.na(MDdf$Age)), "\n")
cat("Predicted Age — Mean:", round(mean(pred_age),1), "| SD:", round(sd(pred_age),1),
    "| Range:", min(pred_age), "-", max(pred_age), "\n")
```

---

### MD Step 7: Multiple Imputation (MICE)

MICE generates multiple complete datasets by imputing each variable conditioned on all others. **Management dictates m=10 imputation cycles.** Methods are assigned automatically by variable type. We extract one completed dataset for modeling; use `with()` + `pool()` for pooled inference.

```{r mice_run, echo=TRUE}
mice_data <- MDdf[, impute_vars]

# Initialize to get default method vector
init <- mice(mice_data, maxit=0, print=FALSE)
meth <- init$method

# Build smarter predictor matrix: only use predictors with >= 10% correlation
# AND >= 50% usable cases (avoids multicollinearity from the 70%-missing block)
pred <- quickpred(mice_data, mincor=0.1, minpuc=0.5)

# Assign methods by variable type for variables still needing imputation
for(v in impute_with_na) {
  cls <- class(mice_data[[v]])[1]
  if(cls %in% c("numeric","integer")) {
    meth[v] <- "pmm"
  } else if(cls == "factor") {
    meth[v] <- if(nlevels(mice_data[[v]]) == 2) "logreg" else "polyreg"
  }
}

cat("Methods assigned:\n")
print(meth[meth != ""])

# Run MICE — MANAGEMENT DICTATE: m=10 imputation cycles
mice_imp <- mice(mice_data, method=meth, predictorMatrix=pred,
                 m=10, maxit=10, seed=42, print=FALSE)
cat("\nMICE complete: m=10 datasets (management dictate), maxit=10 iterations\n")
```

```{r mice_diagnostics, echo=TRUE, fig.width=10, fig.height=6}
# Convergence: chains should mix freely without drift
plot(mice_imp, layout=c(4,3))
```

```{r mice_density, echo=TRUE, fig.width=10, fig.height=6}
# Density: blue = observed, red = imputed (should roughly overlap for numeric vars)
densityplot(mice_imp, ~ DistA + Tenure + Net_Worth_True + HouseHold_Income_True)
```

```{r mice_complete, echo=TRUE}
# Extract completed dataset (imputation #1)
MDdf_imputed <- complete(mice_imp, 1)

remaining_na <- colSums(is.na(MDdf_imputed[, impute_with_na]))
if(sum(remaining_na) == 0) {
  cat("All", length(impute_with_na), "variables fully imputed — 0 NAs remaining.\n")
} else {
  print(remaining_na[remaining_na > 0])
}

cat("Final dataset:", nrow(MDdf_imputed), "rows x", ncol(MDdf_imputed), "cols\n")
tryCatch({
  write.csv(MDdf_imputed, "MDdf_imputed.csv", row.names=FALSE)
  cat("Saved to MDdf_imputed.csv\n")
}, error = function(e) cat("Note: Could not save MDdf_imputed.csv (file may be open):", conditionMessage(e), "\n"))
```

**Observations:**

- **pmm** (predictive mean matching): DistA, Tenure, Net_Worth_True, HouseHold_Income_True — preserves distribution without assuming linearity.
- **logreg**: Team_B_STH, Team_C_STH — binary factors.
- **polyreg**: Marital, Educational_Level, Favorite_Caps_Player, Favorite_Sport, Job_Sector, Mode_Of_Transport — multinomial factors.
- **m=10 (Management Dictate):** 10 imputed datasets generated independently — standard practice is 5–10; management specifies 10 here.
- Convergence plots should show flat, freely mixing chains. Trending lines indicate more iterations may be needed.
- Density plots: imputed (red) should broadly follow observed (blue). Divergence in the 70%-missing block is expected — high caution on those variables.

---

### Summary: MD Steps 1–7

| Step | Action | Key Finding |
|------|--------|-------------|
| 1. Identify | Count & table all missing data | 12.2% overall; 7-variable block at 70% |
| 2. Mark | Create M_ indicator variables | Block pattern confirmed (r = 1.0) |
| 3. Clean | Cap Survey_Comp; keep Marital "U" | 110 values corrected |
| 4. Decide | Define exclude list; keep all variables | No rows or analytical vars excluded |
| 5. Assess | Decision trees on M_ indicators | Mix of MAR, MCAR, MNAR (block) |
| 6. Impute | Median (2), Mode (2), Stochastic (2), Regression (1) | 7 variables resolved; 12 go to MICE |
| 7. MICE | **Mgmt Dictate** m=10; pmm / logreg / polyreg | All variables imputed; saved to MDdf_imputed.csv |

